---
title: "Class 11: Comparative Structure Bionformatics"
author: "Christopher Levinger (A17390693)"
format: html
---

Load up the packages we will need for analysis of protein structure sets.

```{r}
library(bio3d)
```

```{r}
install.packages("BiocManager")
BiocManager::install("msa")
```

We will analyze the ADK family with a single ADK database accession code: "1ake_A"

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```
We can search the PDB database to find all related entries.

```{r}
blast <- blast.pdb(aa)
```

```{r}
attributes(blast)
```
```{r}
head(blast$hit.tbl)
```
Make a little summary figure of these results: 

```{r}
hits <- plot(blast)
```

Our "top hits" i.e the most similar entries in the database are:

```{r}
hits$pdb.id
```

```{r}
#Downoload related pdb files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```
```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

Align and superimpose all these structures: 

Sidenote: 

```{r}
library(bio3dview)

view.pdbs(pdbs)
```

This is better but still difficult to see what is similar and different in all these structures or indeed learn much about how this family works.

Let's try PCA:

```{r}
pc <- pca(pdbs)
plot(pc)
```

```{r}
plot(pc, pc.axes=1:2)
```

```{r}
view.pca(pc)
```

Write a PDB "trajectory" for mol-star

```{r}
mktrj(pc, file="pca_results.pdb")
```

Allow for downloading of results